Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-02-22 response-to-reviews.Rmd

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Dear Editor:

We are pleased to submit a revised version of our manuscript, "Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications" (NPH-TR-2021-38020), for consideration for publication in *New Phytologist*. We have addressed all points raised by the reviewers, as detailed below. The most significant changes include:

- We have brought in an additional author, Thomas N. Buckley, to round out the expertise of the author team and help address some of the reviewer comments (particularly Reviewer 2).

@@

**Unfortunately, we've been unable to find many studies examining vertical gradients at the interspecific level, and have included those of which we are aware.**

I find it interesting to try and link the empirical and theoretical knowledge on vertical gradients in forests with what is done in DGCMs and what should be done to improve their accuracy. However, I think there is a big step between the review (part I and II) and the author’s view on the implications for models (Scaling across space and time). I wonder if it would be possible to link a little bit more the different parts, and maybe without going that much in the technical details (cohort, big leaf, multi layers, ...) which I think are sometimes imprecise or maybe wrong. Would it be possible to explain more simply what are the vertical gradients presented in the section II, that are misrepresented today in DGCMs? Maybe by explaining which of the ‘biophysical variables’ and ‘leaf traits’ that you listed, are considered constant when they should be variable vertically?

**Thank you for the comment. We included a section that explains how vertical gradients are represented in models:**

- **Models partition incoming radiation and radiation within the forest vertical profile (i.e., direct vs. diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018]. Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., CLM), while models with multiple vegetative layers can analytically solve the 2-stream approximation for each layer. Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing LAI, or lower light conditions [e.g., @krinnerDynamicGlobalVegetation2005a].**

@@

Most DGCMs represent a vertical variability in the main physiological parameters (Vcmax, Jmax, Rdark, Na). See for example (Krinner et al., 2005; Clark et al., 2011; Oleson et al., 2013). They all use some form of an exponential decrease in the photosynthetic parameters from the canopy to the ground. Since a lot of parameters and variables depend on Vcmax (for example A and therefore gs), this allows the representation of the vertical variation of a lot of leaf traits and variables. Big leaf models have this capability (see for example Krinner et al. 2005 that you cited, Appendix, or Clark et al. 2011). I think it is not clear in the way you wrote this paragraph, and it looks like ‘big leaf’ models do not represent any vertical variation at all. Note also that it is possible to use multi-layer models with or without considering shaded and sunlit leaves (See for example Clark et al. 2011).

**Thank you for noting the need to clarify this point. We have added a passage at the end of the sentence that mentions big-leaf models to clarify that they do indeed account for assumed vertical profiles: “...to big-leaf models that reduce 3D vegetation structure across the entire biosphere into a single vegetation layer, ***implicitly capturing vertical profiles in light, photosynthetic capacity and other features by assuming those profiles are exponential and thus can be integrated analytically…”***(added text italicized here). We also deleted the clause that occurred two sentences later, that had said about big-leaf models: “...which do not represent any vertical stratification…”**

**In addition, we also added a paragraph that more clearly describes the methods used for light competition that enables the type of variation in physiological traits that you described (e.g., Vcmax):**

- **Owing to differences in the representation of forest vertical strata, DVGMs vary in their ability to incorporate vertical variation in leaf traits and physiological processes. In general, however, this variation is accounted for via light competition. Models partition incoming radiation and radiation within the forest vertical profile (i.e., direct vs. diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018]. Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., CLM), while models with multiple vegetative layers can analytically solve the 2-stream approximation for each layer. Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing LAI, or lower light conditions [e.g., @krinnerDynamicGlobalVegetation2005a].**

@@

**We have revised the first sentence here to read: "Biochemical protection against foliage light and heat damage increase with peak radiation loads, and thus tend to be higher in the upper canopy than in the understory (Table 1)." We note that it is not necessarily obvious that mid- or lower-canopy leaves should have less photoprotection, given that they may experience nearly the same peak radiation loads, during sunflecks, as leaves in the upper canopy, despite generally having less capacity to dissipate light by photochemistry. We also choose to keep this rather obvious sentence because we feel that basic topic sentences are important for helping to guide readers from a diversity of sub-disciplines through the manuscript.**

**We separated VAZ section from VOC, and provided further context to NPQ in the following paragraph:**

- **Biochemical protection against foliage light and heat damage increases with peak radiation loads, and thus tends to be higher in the upper canopy than in the understory (Table 1). Further, by reducing photosynthesis, heat and VPD-associated stomatal closure can indirectly reduce the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007] <!--Nidhi, I feel like this sentence needs a reference. {maybe this would be better as a topic sentence without a ref? but there is one Ty suggested-->. Accumulation of excess light energy results in overreduction of the electron transport chain and the formation of harmful reactive oxygen species [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a]. A ubiquitous defense is a rapidly inducible non-photochemical quenching (NPQ) mechanism that responds to the increased thylakoid pH gradient caused by accumulated light energy [@niyogiSafetyValvesPhotosynthesis2000; @gossBiodiversityNPQ2015]. This form of NPQ is accomplished by the interconversion of xanthophyll cycle pigments—violaxanthin, antheraxanthin, and zeaxanthin (VAZ)—which regulates the capacity for de-excitation of chlorophyll through thermal dissipation instead of photochemistry. Due to higher leaf temperatures and hydraulic limitations, upper canopy leaves spend more time with stomata closed (Table 2), and therefore require a high capacity to dissipate excess light energy [@niinemetsPhotosynthesisResourceDistribution2007]. Correspondingly, leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table 1, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a]. In contrast, in order to maximize carbon uptake and minimize damage during $T_{leaf}$ surges caused by sunflecks (see section 2.2), shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light compared to sun-acclimated, upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].**

@@

There are a few dense sentences or run-ons that could be revised. Lines 192-196 and lines 201-204, for instance. Please revise.

**We revised the following lines:**

- **Lines 192-196: Notably, this pattern can be reversed in open forests, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to a ‘canopy greenhouse effect’**

- **Notably, buffering against high maximum daytime $T_{air}$ becomes weaker and can be reversed as canopies become more open. Analogous to a 'canopy greenhouse effect', in open forests, below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$. This is due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers**

@@

- **Previous sentences: Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012]. Likewise, the mid-canopy experiences a highly variable light environment due to the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].**

- **Changed sentences: Mid-canopies and understories experience a highly dynamic light environment due to sunflecks, or brief increases in direct solar radiation or PAR, caused by small canopy gaps, wind-induced canopy movements or the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].**

Fig. 1: In the diagram, stomatal density, leaf angle are separate categories yet VAZ and VOC are not. As far as I know though, there is no biochemical link between VAZ and VOC, so these should be separate shouldn't they?

**We agree with the suggestions. Therefore, to make it more clearer, we added photoprotection as one of the factors in Fig.1 and isoprene emission in a separate line instead of VAZ and VOC in the same line.**

@@

Fig. 2 is a mix of computed and measured quantities. LAD and proportion of sun leaves are both computed quantities, and the basis of these should be clearer. Or at least state 'modelled LAD' and 'calculated proportion of sun leaves'.

**Done**

line165-172: refixation of respired CO2 in the understory isn’t mentioned here, but there have been a number of studies, particularly involving 13C, that have looked at this. Please mention along with at least 1 reference.

**Vertical CO~2~ concentration gradients can influence the ratio of leaf intercellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$) calculated from $\delta^{13}C$, and refixation of respired CO~2~ vertically, where both tend to be higher at lower heights (Table 1). Particularly, understory seedlings benefit from elevated CO~2~ by having highest rates of CO~2~ refixation [@brooksVerticalGradientsPhotosynthetic1997]."**

@@

line187: Where mentioning radiation fluxes, could I please ask the authors to use the phrase 'and sensible heat emission'? I think this is really what the authors are referring to, so why not say so?

**Done**

line188: 'dense canopies buffer understories' is one perspective, that there is far less radiant energy received and therefore understories stay cool is another. Can the authors please consider reflecting this perspective? I’ve studies canopies for a a score of years and I just don’t think 'buffering' is at play so much!

**We were using the term “buffer” in a functional or biological sense, meaning “protect,” rather than as a description of the underlying physics; i.e., the canopy protects the understory from large diel swings in both air temperature and radiation. We have reworded this passage to clarify: “Typically, ***diel temperature range is smaller beneath dense canopies than above, resulting in,***~~dense canopies buffer understories from high maximum $T_{air}$ more than open canopies, i.e., dense canopy understories can have~~ cooler maximum daytime $T_{air}$ and warmer nighttime minimum Tair than open canopy understories or nearby clearings.”**

@@

line206-207 'Leaves are typically warmer than air … '. That is not what is predicted in theory, proven by parts of the simulations in Fig. 3. Why not use 'Leaves can be warmer …'? That would be more consistent.

**Done**

Figure 3: Variables in the inset table need to be stated. I can guess what they are, but this should be explicit.

**We added the description for the variables as part of the figure caption: Variables in the inset table are now described in the figure caption as: "Biohphysical constants in the the table include: shortwave radiation (swr), wind speed (ws), relative humidity (rh), leaf characteristic dimension (lcd), stomatal conductance (gs), air temperature (tair)."**

@@

Table 1: What is here is not leaf gas exchange, and it seems that the authors have repeated the caption for Table 2 in error. Please fix this. Also don’t capitalise ‘leaf’.

**Done**

Lines 354-372 on sun versus shade leaves and Table 1. Most of what is here was known a long time ago, say in Vogel 1968 (cited) and Boardman 1977 (not cited) but pertains to within-canopy rather than between sun and shade plants. I would suggest that this would be improved if the authors started from these old studies that most readers will recognise, and state how we’ve advanced in understanding and progressed toward within-canopy work rather than between sun-and-shade habitats.

**Thank you for the suggestion. We have added additional information highlighting older work and recent advances in the paragraph below:**

**Leaf traits vary vertically across forest strata at anatomical, structural and biochemical levels [Table 1, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping $T_{leaf}$, leaf metabolism, and leaf thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a; @zwienieckiHydraulicLimitationsImposed2004]. This vertical variation in leaf structure and composition has been long recognized, with sun and shade leaf traits distinguished for more than 100 years [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928]. Early work attributed differences among leaves to sun versus shade [@wyliePrinciplesFoliarOrganization1951; @vogelSunLeavesShade1968], and this tendency grew given the parallel differences observed for plants grown in sun versus shade conditions [@boardmanComparativePhotosynthesisSun1977; @abramsLeafStructuralCharacteristics1990]. More recent work has included a focus not only on light, but also on leaf height in the canopy, as important in determining differences in mature leaf structure and function, as leaf traits may partially mitigate the effects of gravity and hydraulic pathlength that would affect leaves higher up in the canopy [@sackHowStrongIntracanopy2006; @kochLimitsTreeHeight2004; @burgessRegressionsLeafTraits2006]. For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022]. Thus, leaves develop differently throughout the canopy according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of the ontogenetic stage of the branches [Table 1, @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015]. Plasticity throughout the canopy may emerge during new leaf development with leaves being influenced by information from adjacent mature leaves or apical meristem, including  (e.g., for deciduous species expanding leaves) a “memory” of previous conditions in that location, and may also acclimate during and after expansion to to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004]. Overall, leaves may optimize their biochemical, anatomical and structural traits according to their light environment and height [@lloydOptimisationPhotosyntheticCarbon2010; @buckleyOptimalCarbonPartitioning2021; @hikosakaOptimalNitrogenDistribution2014; @kitaoCanopyNitrogenDistribution2018; @niinemetsPhotosynthesisResourceDistribution2007]. Yet, while these studies have clarified many details of leaf differences throughout the canopy, no papers have integrated the many differences in leaf traits throughout the canopy in a cohesive framework. Rather, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species using sun leaves [@wrightWorldwideLeafEconomics2004; @keenanGlobalLeafTrait2016; @chenLeafEconomicsSpectrum2020]. As canopy shade is known to alter these trait relationships [@osnasDivergentDriversLeaf2018], further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.**

@@

line485-487: It seems that VPD and its effect on gs is ignored here. Please clarify? The well-known stomatal closure at high VPD alone would mean higher Tair would decrease gs in the upper canopy (not 'canopy leaves' as stated) compared to the lower canopy. See Grossiord et al. (2020) Tansley Review. VPD only gets mentioned in the section on photosynthesis (l. 496). Please do so earlier.

**We agree, and have modified the text in this section to emphasize the role of VPD in driving stomatal closure in sun leaves (new text italicized): “However, ~~because~~ water supply often cannot meet the demands incurred by the high irradiance and $g_b$ experienced by sun leaves ***with fully open stomata, because of height-related constraints on water transport (e.g., Yoder et al. 1994, Koch et al. 2004, Sillett et al 2010) and/or because of increased evaporative demand (VPD) caused by leaf warming in sunlit canopy locations (Buckley et al. 2014)***. As a result, midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests (Table 2), which drives the lower intracellular CO2 and carbon isotope discrimination discussed in section 3.1 (Table 1).  Stomatal depression reduces transpirational cooling, thus amplifying the warming of sun leaves by high radiation loads ~~In sun leaves, Tleaf thus further increases due to the lack of transpirational cooling.~~” We have also added the phrase ***“(driven by increased VPD)”***after “sun leaves in the upper canopy show a stronger decrease in gs in response to rising Tair”, and clarified that “high Tair should decrease gs of ***upper canopy***leaves”.**

line510-525: There is a long discussion about whether Topt for gas exchange or its components differs between overstory and understory. Here the authors aren’t clear, but such a comparison only matters if the same species is measured in different canopy layers.

**Thank you for noting this, we revised the sentence to make it clear (please see below), where we specify that majority of the studies are within species between sun and shade leaves of the same trees. @carterExperimentalWarmingTropical2021 is one of the only studies comparing between overstory and understory.**

### 2022-02-22 response-to-reviews.Rmd

**Regarding incorporation of more theory, please see our response to the comment below.**

**We have made the suggested improvements to figure 2: lines are bolder, and height normalized to height at the top of the canopy.**

Certainly, a discussion with a micrometeorologist or atmospheric scientist, if not already had, could improve the manuscript. Statements like l. 156: 'Wind speeds are also higher at the top of the canopy, owing to the buffering effect of the canopy' are an awkward read. Buffering? There could be much improvement by looking at it the other way: wind will blow as it does at the top of the canopy until encountering the plant canopy top as an aerodynamic drag element; then the additional leaf area entrains air movement and sweep-eject motion and eddies are attenuated through the canopy resulting in progressively lower windspeeds deeper into canopies. I don’t know how buffering enters into this, or what the authors mean here.

**We appreciate these comments and have added a discussion of canopy transport from a more theoretical perspective (lines 169-192), immediately after the paragraph about canopy radiation profiles and before those on CO~2~, H~2~O and temperature profiles. We have also modified the subsequent paragraph about wind speeds to eliminate the use of the confusing term “buffering”.**

@@

**We had also used the words “buffering”, "buffer", or "buffered" in a different context, where the intention was to describe how canopy attenuation (of radiation and wind) can reduce daily extremes of temperature in the lower canopy and understory. We have modified the text in those places to clarify, generally reducing use of the terms. However, the term has been widely used in recent publications on this theme, and we therefore did not feel it was necessary to eliminate it's usage. Examples of modified sentences include the following:**

- **“Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions ***from large fluctuations in conditions above the canopy”; (line 127)***(new text italicized)**
- **“It is most commonly the case that attenuation of radiation and vertical transport of sensible heat by the canopy buffers the lower canopy and understory from large diel and seasonal swings in air temperature" (line 229)**
- **“Typically, ***diurnal temperature range is smaller beneath dense canopies than above, resulting in,***~~dense canopies buffer understories from high maximum $T_{air}$ more than open canopies, i.e., dense canopy understories can have~~cooler maximum daytime $T_{air}$ and warmer nighttime minimum Tair than open canopy understories or nearby clearings.” (line 232)**

### 2022-02-22 vertical-thermal-review.Rmd

Text | word count | other | n
---|---|-----|--------------------
Total word count (excluding summary, references and legends) | 	9559 |	No. of figures | 	6 (all colour)
Summary|  	200 | 	No. of Tables | 	2  
Introduction|	391 |	No of Supporting Information files | 	6
Review of vertical gradients |	6484	||	 
Implications |	2408	 ||
Conclusions |	276	||
Acknowledgments | 43	||

>>>>

Text | word count | other | n
---|---|-----|--------------------
Total word count (excluding summary, references and legends) | 	9495 |	No. of figures | 	6 (all colour)
Summary|  	200 | 	No. of Tables | 	2  
Introduction|	393 |	No of Supporting Information files | 	6
Review of vertical gradients |	6374	||	 
Implications |	2408	 ||
Conclusions |	276	||
Acknowledgments | 43	||
